---
title: "Hands-on Exercise 4c: Visualising Uncertainty"
author: "Lei Yuhe"
date: "May 10, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Overview

Raw summary statistics (e.g. a mean value) are **point estimates**, they tell us nothing about how confident we should be in that number. Communicating uncertainty helps us make better decisions and avoids misleading conclusions.

These will be covered below: 

- Plot **standard error bars** using `ggplot2`. 
- Plot **confidence interval** error bars using `ggplot2`. 
- Create **interactive error bars** by combining `ggplot2`, `plotly`, and `DT`. 
- Build advanced uncertainty visuals using the **`ggdist`** package. 
- Create **Hypothetical Outcome Plots (HOPs)** using the `ungeviz` package.

---

# Setup

## Key Terms

| Term | What it measures | Formula |
|---|---|---|
| **Standard Deviation (SD)** | Spread of individual observations | $s$ |
| **Standard Error (SE)** | Uncertainty of the sample *mean* | $SE = s / \sqrt{n-1}$ |
| **Confidence Interval (CI)** | Plausible range for the true mean | $\bar{x} \pm z \times SE$ |

> **Important:** Do **not** confuse the uncertainty of a point estimate (SE / CI)
> with the variation in the sample (SD). They answer different questions.

## Installing and Loading Packages

```{r}
#| label: load-packages

pacman::p_load(
  plotly,       # Interactive plots
  crosstalk,    # Cross-widget interactions (linked brushing/filtering)
  DT,           # Interactive HTML tables
  ggdist,       # Distribution and uncertainty visualisation
  ggridges,     # Ridgeline plots
  colorspace,   # Colour palette utilities
  gganimate,    # Animated plots
  tidyverse     # Core data science packages (ggplot2, dplyr, etc.)
)
```

## Data Import

We use `Exam_data.csv`, which contains exam scores (Maths, English, Science) for
primary school students, along with demographic variables such as `RACE`, `GENDER`,
and `CLASS`.

```{r}
#| label: import-data

exam <- read_csv("data/Exam_data.csv")
glimpse(exam)
```

---

# Visualising Uncertainty with ggplot2

## Deriving Summary Statistics

To plot error bars we first need to compute, per group:

- **n**: number of observations
- **mean**: average Maths score
- **sd**: standard deviation
- **se**: standard error = $sd / \sqrt{n-1}$

```{r}
#| label: summary-stats

my_sum <- exam %>%
  group_by(RACE) %>%
  summarise(
    n    = n(),
    mean = mean(MATHS),
    sd   = sd(MATHS)
  ) %>%
  mutate(se = sd / sqrt(n - 1))

knitr::kable(my_sum, format = "html", digits = 2,
             caption = "Summary statistics of Maths scores by Race")
```

**Reading the table:**

- Chinese pupils have the largest group ($n = 193$) and the smallest SE ($\approx 1.13$),
  which means we have a precise estimate of their mean.
- Indian pupils have the smallest group ($n = 12$) and a very large SE ($\approx 7.04$),
  reflecting much greater uncertainty about their true mean.

---

## Standard Error Bars

Standard error bars extend one SE above and below the mean. They communicate
**how precisely** the sample mean estimates the population mean. Narrower bars
mean more precision.

```{r}
#| label: fig-se-bars
#| fig-cap: "Standard error bars of mean Maths score by race. Red dots = sample means; bars = ±1 SE."

ggplot(my_sum) +
  geom_errorbar(
    aes(x = RACE, ymin = mean - se, ymax = mean + se),
    width     = 0.2,
    colour    = "black",
    alpha     = 0.9,
    linewidth = 0.5
  ) +
  geom_point(
    aes(x = RACE, y = mean),
    stat  = "identity",
    color = "red",
    size  = 1.5,
    alpha = 1
  ) +
  labs(
    title = "Standard Error of Mean Maths Score by Race",
    x     = "Race",
    y     = "Mean Maths Score"
  ) +
  theme_minimal()
```

**What to notice:**

- The Chinese group has a very tight error bar (large sample, high precision).
- The Indian and Others groups have much wider bars (small samples, low precision).
- The visual inequality in bar widths reinforces why sample size matters when
  comparing groups.
  
---

## 95% Confidence Interval Bars

A 95% CI is computed as $\bar{x} \pm 1.96 \times SE$. It means: if we repeated
this study many times, ~95% of the constructed intervals would contain the true
population mean.

The races are **reordered by descending mean** so the plot is easier to read.

```{r}
#| label: fig-ci95-bars
#| fig-cap: "95% confidence intervals of mean Maths score by race, sorted by descending mean."

ggplot(my_sum) +
  geom_errorbar(
    aes(
      x    = reorder(RACE, -mean),
      ymin = mean - 1.96 * se,
      ymax = mean + 1.96 * se
    ),
    width     = 0.2,
    colour    = "black",
    alpha     = 0.9,
    linewidth = 0.5
  ) +
  geom_point(
    aes(x = reorder(RACE, -mean), y = mean),
    stat  = "identity",
    color = "red",
    size  = 1.5,
    alpha = 1
  ) +
  labs(
    x     = "Race",
    y     = "Mean Maths Score",
    title = "95% Confidence Interval of Mean Maths Score by Race"
  ) +
  theme_minimal()
```

**Key difference from SE bars:** The multiplier changes the interval width:

| Confidence Level | Multiplier | Interpretation |
|---|---|---|
| ~68% | ±1.00 SE | Equivalent to ±1 SE bars |
| 95%  | ±1.96 SE | Standard CI in most research |
| 99%  | ±2.58 SE | More conservative, wider bars |

## Interactive Error Bars (99% CI)

By combining `plotly` (interactive chart), `DT` (interactive table), and
`crosstalk` (linked selection), we can build a dashboard where clicking a row in
the table highlights the corresponding bar in the plot.

```{r}
shared_df = SharedData$new(my_sum)

bscols(widths = c(4,8),
       ggplotly((ggplot(shared_df) +
                   geom_errorbar(aes(
                     x=reorder(RACE, -mean),
                     ymin=mean-2.58*se, 
                     ymax=mean+2.58*se), 
                     width=0.2, 
                     colour="black", 
                     alpha=0.9, 
                     size=0.5) +
                   geom_point(aes(
                     x=RACE, 
                     y=mean, 
                     text = paste("Race:", `RACE`, 
                                  "<br>N:", `n`,
                                  "<br>Avg. Scores:", round(mean, digits = 2),
                                  "<br>95% CI:[", 
                                  round((mean-2.58*se), digits = 2), ",",
                                  round((mean+2.58*se), digits = 2),"]")),
                     stat="identity", 
                     color="red", 
                     size = 1.5, 
                     alpha=1) + 
                   xlab("Race") + 
                   ylab("Average Scores") + 
                   theme_minimal() + 
                   theme(axis.text.x = element_text(
                     angle = 45, vjust = 0.5, hjust=1)) +
                   ggtitle("99% Confidence interval of average /<br>maths scores by race")), 
                tooltip = "text"), 
       DT::datatable(shared_df, 
                     rownames = FALSE, 
                     class="compact", 
                     width="100%", 
                     options = list(pageLength = 10,
                                    scrollX=T), 
                     colnames = c("No. of pupils", 
                                  "Avg Scores",
                                  "Std Dev",
                                  "Std Error")) %>%
         formatRound(columns=c('mean', 'sd', 'se'),
                     digits=2))

```

**Why interactivity helps:** Hovering over a point reveals exact values that
are hard to read off a static plot. The linked table lets you filter by group
without redrawing the chart.

# Visualising Uncertainty with ggdist

The `ggdist` package extends `ggplot2` with geoms and stats purpose-built for
distribution and uncertainty visualisation. It works seamlessly for both
frequentist (CI / SE) and Bayesian (posterior distribution) contexts.

**Core philosophy:** uncertainty *is* a distribution — so draw it like one.

## `stat_pointinterval()` Point + Interval

`stat_pointinterval()` shows a point estimate (default: mean) together with
multiple nested intervals (default: 66% and 95%).

```{r}
#| label: fig-pointinterval-default
#| fig-cap: "Mean and multiple confidence intervals of Maths scores by race using stat_pointinterval()."

exam %>%
  ggplot(aes(x = RACE, y = MATHS)) +
  stat_pointinterval() +
  labs(
    title    = "Visualising Confidence Intervals of Mean Maths Score",
    subtitle = "Mean Point + Multiple-interval plot",
    x        = "Race",
    y        = "Maths Score"
  ) +
  theme_minimal()
```

The thick inner line = 66% CI; the thin outer line = 95% CI. This **nested
interval** design lets you read two confidence levels at once.

---

## `stat_pointinterval()` Median + Quantile Interval

We can switch to the **median** as the point summary and use **quantile intervals
(qi)** instead of the default. This is more appropriate for skewed distributions.

```{r}
#| label: fig-pointinterval-median
#| fig-cap: "Median and 95% quantile interval of Maths scores by race."

exam %>%
  ggplot(aes(x = RACE, y = MATHS)) +
  stat_pointinterval(
    .width    = 0.95,
    .point    = median,
    .interval = qi
  ) +
  labs(
    title    = "Visualising Confidence Intervals of Median Maths Score",
    subtitle = "Median Point + 95% Quantile Interval",
    x        = "Race",
    y        = "Maths Score"
  ) +
  theme_minimal()
```

**When to use median + qi:**

- When your data may be skewed (median is more robust than mean).
- When you want a distribution-free summary rather than a Normal-theory CI.

---

## `stat_pointinterval()` Multiple Widths (95% & 99%)

By passing a vector to `.width`, we can display both 95% and 99% CIs
simultaneously, encoded by line thickness.

```{r}
#| label: fig-pointinterval-multi
#| fig-cap: "Mean with 95% and 99% confidence intervals shown simultaneously."

exam %>%
  ggplot(aes(x = RACE, y = MATHS)) +
  stat_pointinterval(
    .width      = c(0.95, 0.99),
    show.legend = TRUE
  ) +
  labs(
    title    = "95% and 99% Confidence Intervals of Mean Maths Score",
    subtitle = "Mean Point + Multiple-interval plot",
    x        = "Race",
    y        = "Maths Score"
  ) +
  theme_minimal()
```

The thicker inner interval = 95% CI; the thinner outer interval = 99% CI.
The 99% interval is always wider, as we need a broader net to be more confident
about catching the true mean.

---

## `stat_gradientinterval()` Gradient Uncertainty

`stat_gradientinterval()` encodes probability density through **colour opacity**:
regions of higher likelihood are darker/more opaque, fading toward the tails.
This gives a more intuitive sense of where values are most plausible.

```{r}
#| label: fig-gradientinterval
#| fig-cap: "Gradient interval plot: opacity encodes density of plausible values."

exam %>%
  ggplot(aes(x = RACE, y = MATHS)) +
  stat_gradientinterval(
    fill         = "skyblue",
    show.legend  = TRUE
  ) +
  labs(
    title    = "Visualising Confidence Intervals of Mean Maths Score",
    subtitle = "Gradient + Interval Plot",
    x        = "Race",
    y        = "Maths Score"
  ) +
  theme_minimal()
```

**Why gradient plots are powerful:**

- They avoid the false binary of "inside the interval = plausible, outside =
  implausible".
- The continuous fade communicates that probability *decreases gradually* toward
  the tails.
- Particularly useful when communicating with non-statistical audiences.

---

# Hypothetical Outcome Plots (HOPs)

## What are HOPs?

A **Hypothetical Outcome Plot** (HOP) is an animated visualisation that cycles
through a set of plausible outcomes drawn from the uncertainty distribution. Rather
than showing a static summary (a bar, an interval), it lets the viewer *experience*
the uncertainty through repeated random draws.

**Cognitive advantage:** Research (Hullman et al., 2015; Kale et al., 2019) shows
that people reason better about uncertainty when they see animated samples than
when they see static interval representations.

## Installing the `ungeviz` Package

The `ungeviz` package is not on CRAN and must be installed from GitHub. You only
need to do this once.

```{r}
#| label: install-ungeviz
#| eval: false

devtools::install_github("wilkelab/ungeviz")
```

```{r}
#| label: load-ungeviz

library(ungeviz)
```


## Building the HOP

The key functions are:

- `sampler(k, group)`: draws `k` bootstrap samples within each group
- `geom_hpline()`: draws a short horizontal line at the sample mean for each draw
- `transition_states(.draw, ...)`: animates through draws using `gganimate`


```{r}
#| label: fig-hops
#| fig-cap: "Hypothetical Outcome Plot (HOP): each frame shows a bootstrap sample mean per race group. The animation conveys how much the mean could plausibly vary."

ggplot(data = exam, 
       (aes(x = factor(RACE), 
            y = MATHS))) +
  geom_point(position = position_jitter(
    height = 0.3, 
    width = 0.05), 
    size = 0.4, 
    color = "#0072B2", 
    alpha = 1/2) +
  geom_hpline(data = sampler(25, 
                             group = RACE), 
              height = 0.6, 
              color = "#D55E00") +
  theme_bw() + 
  transition_states(.draw, 1, 3)
```


**Reading a HOP:**

- Each frame shows one possible sample mean (orange horizontal line) for each race
  group.
- Groups with high uncertainty (small $n$, high SD) will show orange lines that
  jump around a lot between frames.
- Groups with low uncertainty will show stable orange lines.
- This directly shows the viewer that "the mean could plausibly be anywhere in
  this range" rather than asking them to interpret an abstract interval statistic.

---

# Summary

| Method | Package | Best For |
|---|---|---|
| `geom_errorbar()` ± SE | `ggplot2` | Simple, traditional SE bars |
| `geom_errorbar()` ± 1.96·SE | `ggplot2` | Classic 95% CI bars |
| `ggplotly()` + `DT` + `SharedData` | `plotly`, `crosstalk`, `DT` | Interactive exploration |
| `stat_pointinterval()` | `ggdist` | Multiple nested CIs at once |
| `stat_gradientinterval()` | `ggdist` | Continuous density-based uncertainty |
| `geom_hpline()` + `transition_states()` | `ungeviz`, `gganimate` | Animated HOPs for intuitive communication |

**Core takeaway:** Different audiences need different representations.
Traditional SE bars work for statistical experts; gradient intervals and HOPs
are more intuitive for general audiences and improve risk communication.

---


# Reference 

Kam T. S. (2023) [R for Visual Analytics](https://r4va.netlify.app/chap11)










