---
title: "Hands-on Exercise 4"
author: "Lei Yuhe"
date: "May 3, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Visualising Distribution

## Overview

Two new statistical graphic methods for visualizing distribution will be covered in this section. 

1. **Ridgeline plots**: Show distributions for multiple groups stacked along a shared axis, using `ggridges`. 
2. **Raincloud plots**: Combine a half-density, boxplot, and dot plot into one layered graphic, using `ggdist`. 

Both reveal richer distributional shape such as **skewness**, **multimodality**, and **tail behaviour**.

---

## Setting Up

```{r}
#| label: setup
pacman::p_load(ggdist, ggridges, ggthemes,
               colorspace, tidyverse)
```

| Package | Role |
|---|---|
| `ggridges` | Ridgeline plots via `geom_density_ridges()` and variants |
| `ggdist` | Half-eye, dot plots, uncertainty visualisation |
| `ggthemes` | Extra themes, e.g. `theme_economist()` for publication-ready look |
| `colorspace` | Colour manipulation via `lighten()`, `darken()`, palette tools |
| `tidyverse` | Data wrangling (`dplyr`, `tidyr`), and base plotting (`ggplot2`) |

## Importing Data

```{r}
#| label: data-import
exam <- read_csv("data/Exam_data.csv")
glimpse(exam)
```

The dataset has **322 students** across 9 classes (3A–3I) with scores in
English, Maths, and Science. Variables `CLASS`, `GENDER`, and `RACE` serve
as grouping factors throughout this chapter.

---

## Visualising Distribution with Ridgeline Plot
A [**ridgeline plot**](https://www.data-to-viz.com/graph/ridgeline.html) (formerly called a Joyplot) shows the distribution of a numeric variable for several groups, stacked with a slight overlap along the y-axis.

> **When to use ridgeline plots:**
>
> - **5 or more groups** to compare (if we have fewer groups, use violin/boxplot instead)
> - Existence of **natural ordering** to the groups (e.g. class ranking)
> - Reveal **shape/distribution differences** across groups, not just means

In the Exam dataset used here, 9 classes (3A–3I) are a perfect candidate.

### Basic Ridgeline Plot

Distribution of **English scores** across all 9 classes using `geom_density_ridges()`. The classes are ordered 3A to 3I (best to weakest by class assignment).

:::: panel-tabset

#### The plot
```{r}
#| label: fig-ridgeline-basic
#| fig-cap: "English score distributions by class - basic ridgeline plot."
#| fig-height: 4
#| echo: false
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS)) +
  geom_density_ridges(
    scale          = 3,
    rel_min_height = 0.01,
    bandwidth      = 3.4,
    fill           = lighten("#7097BB", .3),
    color          = "white"
  ) +
  scale_x_continuous(
    name   = "English Score",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name   = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_ridges()
```

#### The code
```{r}
#| eval: false
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS)) +
  geom_density_ridges(
    scale          = 3,
    rel_min_height = 0.01,
    bandwidth      = 3.4,
    fill           = lighten("#7097BB", .3),
    color          = "white"
  ) +
  scale_x_continuous(
    name   = "English Score",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name   = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_ridges()
```
::::

**Key arguments to understand:**

- `scale = 3`: It controls how much ridges overlap (higher = more overlap)
- `rel_min_height = 0.01`: It trims tails below 1% of the peak height, avoiding long flat tails
- `bandwidth = 3.4`: It controls the kernel density smoother width. The higher, the smoother. 

> **Takeaway:**
> Class 3A and 3B have distributions clustered at high scores (70-90), while 3H
> and 3I are concentrated at low scores (20-50). Notably, the *shape* differs as well, 
> 3I shows a narrow, peaked distribution (homogeneously weak), while middle
> classes like 3E-3F show broader, flatter distributions indicating mixed ability.

### Varying fill colors along the x axis

Instead of a single flat fill colour, we can let the fill **vary along the score
axis** using `geom_density_ridges_gradient()`. This adds a visual cue for score
magnitude.

:::: panel-tabset
#### The plot
```{r}
#| label: fig-ridgeline-gradient
#| fig-cap: "Maths score distributions with fill colour varying along the score axis."
#| fig-height: 4
#| echo: false

ggplot(exam,
       aes(x    = MATHS,
           y    = CLASS,
           fill = after_stat(x))) +
  geom_density_ridges_gradient(
    scale          = 3,
    rel_min_height = 0.01
  ) +
  scale_fill_viridis_c(
    name   = "Maths Score",
    option = "C"
  ) +
  scale_x_continuous(
    name   = "Maths Score",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name   = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_ridges()
```

#### The code
```{r}
#| eval: false
ggplot(exam,
       aes(x    = MATHS,
           y    = CLASS,
           fill = after_stat(x))) +
  geom_density_ridges_gradient(
    scale          = 3,
    rel_min_height = 0.01
  ) +
  scale_fill_viridis_c(
    name   = "Maths Score",
    option = "C"
  ) +
  scale_x_continuous(
    name   = "Maths Score",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name   = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_ridges()
```
::::

> **Takeaway:**
> The gradient fill immediately communicates which classes cluster at high vs low
> scores without needing to read axis values carefully. However, `fill` and
> `alpha` transparency **cannot be used together** with gradient fills, which is a
> technical limitation to take note. 

> **Code:** `after_stat(x)` is the modern replacement for the deprecated
> `stat(x)` syntax used in the course notes. Both map the computed x-axis value
> onto the fill aesthetic.


### Mapping the probabilities directly onto colour
Beside providing additional `geom` objects to support the need to plot ridgeline plot, `ggridges` package also provides a stat function called [`stat_density_ridges()`](https://wilkelab.org/ggridges/reference/stat_density_ridges.html) that replaces [`stat_density()`](https://ggplot2.tidyverse.org/reference/geom_density.html) of ggplot2.

:::: panel-tabset
#### The plot
```{r}
#| label: fig-ridgeline-ecdf
#| fig-cap: "Science score ridgelines coloured by tail probability (ECDF)."
#| fig-height: 4
#| echo: false

ggplot(exam,
       aes(x    = SCIENCE,
           y    = CLASS,
           fill = 0.5 - abs(0.5 - after_stat(ecdf)))) +
  stat_density_ridges(
    geom      = "density_ridges_gradient",
    calc_ecdf = TRUE
  ) +
  scale_fill_viridis_c(
    name      = "Tail\nProbability",
    direction = -1
  ) +
  scale_x_continuous(name = "Science Score", expand = c(0, 0)) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```
#### The code
```{r}
#| eval: false

ggplot(exam,
       aes(x    = SCIENCE,
           y    = CLASS,
           fill = 0.5 - abs(0.5 - after_stat(ecdf)))) +
  stat_density_ridges(
    geom      = "density_ridges_gradient",
    calc_ecdf = TRUE
  ) +
  scale_fill_viridis_c(
    name      = "Tail\nProbability",
    direction = -1
  ) +
  scale_x_continuous(name = "Science Score", expand = c(0, 0)) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```
::::

> **Takeaway:**
> The expression `0.5 - abs(0.5 - ecdf)` maps values **closest to the median**
> to the highest fill value, and tail extremes to the lowest. Dark regions are
> the rare extreme scores. Bright central bands show where most students fall.
> This is especially useful for spotting which classes have fat tails (more
> outliers) vs tight central distributions.

::: callout-important
Always include `calc_ecdf = TRUE` inside `stat_density_ridges()` when using ECDF-based colour mapping, the plot will silently produce incorrect results without it.
:::

### Ridgeline plots with quantile lines
We can colour the ridgeline by **quartile bands** to make Q1, median, and Q3
immediately visible within each class's density shape. 

:::: panel-tabset 
#### Quartile Band 
``` {r}
#| label: fig-ridgeline-quantile
#| fig-cap: "English score ridgelines coloured by quartile band."
#| fig-height: 4
#| echo: false

ggplot(exam,
       aes(x    = ENGLISH,
           y    = CLASS,
           fill = factor(after_stat(quantile)))) +
  stat_density_ridges(
    geom           = "density_ridges_gradient",
    calc_ecdf      = TRUE,
    quantiles      = 4,
    quantile_lines = TRUE
  ) +
  scale_fill_viridis_d(name = "Quartiles") +
  scale_x_continuous(name = "English Score", expand = c(0, 0)) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

#### The code
``` {r}
#| eval: false

ggplot(exam,
       aes(x    = ENGLISH,
           y    = CLASS,
           fill = factor(after_stat(quantile)))) +
  stat_density_ridges(
    geom           = "density_ridges_gradient",
    calc_ecdf      = TRUE,
    quantiles      = 4,
    quantile_lines = TRUE
  ) +
  scale_fill_viridis_d(name = "Quartiles") +
  scale_x_continuous(name = "English Score", expand = c(0, 0)) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```


#### Probability Band
``` {r}
#| label: fig-ridgeline-tails
#| fig-cap: "English scores with 2.5% and 97.5% tail regions highlighted in red/blue."
#| fig-height: 4
#| echo: false

ggplot(exam,
       aes(x    = ENGLISH,
           y    = CLASS,
           fill = factor(after_stat(quantile)))) +
  stat_density_ridges(
    geom      = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = c(0.025, 0.975)
  ) +
  scale_fill_manual(
    name   = "Probability",
    values = c("#FF0000A0", "#A0A0A0A0", "#0000FFA0"),
    labels = c("Bottom 2.5%", "Middle 95%", "Top 2.5%")
  ) +
  scale_x_continuous(name = "English Score", expand = c(0, 0)) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

#### The code
``` {r}
#| eval: false

ggplot(exam,
       aes(x    = ENGLISH,
           y    = CLASS,
           fill = factor(after_stat(quantile)))) +
  stat_density_ridges(
    geom      = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = c(0.025, 0.975)
  ) +
  scale_fill_manual(
    name   = "Probability",
    values = c("#FF0000A0", "#A0A0A0A0", "#0000FFA0"),
    labels = c("Bottom 2.5%", "Middle 95%", "Top 2.5%")
  ) +
  scale_x_continuous(name = "English Score", expand = c(0, 0)) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

::::



