---
title: "Hands-on Exercise 4d: Building Funnel Plot with R"
author: "Lei Yuhe"
date: "May 3, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Funnel Plots for Fair Comparison

## Overview

Three methods for creating **funnel plots** for fair visual comparison:

1.  The `FunnelPlotR` package (quick method)
2.  `ggplot2` from scratch (manual method)
3.  `plotly` + `ggplot2` for interactivity

The dataset used is **COVID-19_DKI_Jakarta.csv**, which contains cumulative COVID-19 cases and deaths by sub-district (*kelurahan*) as at 31 July 2021.

------------------------------------------------------------------------

## Load Packages

```{r}
#| label: load-packages

pacman::p_load(tidyverse, FunnelPlotR, plotly, knitr)
```

**Packages used:**

| Package       | Purpose                                  |
|---------------|------------------------------------------|
| `tidyverse`   | Data wrangling & `ggplot2` visualisation |
| `FunnelPlotR` | High-level funnel plot creation          |
| `plotly`      | Interactive visualisation                |
| `knitr`       | Rendering static HTML tables             |

------------------------------------------------------------------------

## Import Data

```{r}
#| label: import-data

covid19 <- read_csv("data/COVID-19_DKI_Jakarta.csv") %>%
  mutate_if(is.character, as.factor)
```

::: callout-note
The dataset covers 267 sub-districts across five cities in DKI Jakarta. Key columns are `Positive` (cumulative confirmed cases) and `Death` (cumulative deaths).
:::

Preview the first few rows:

```{r}
#| label: preview-data

covid19 %>%
  head(6) %>%
  kable()
```

## FunnelPlotR Methods

### Basic Funnel Plot

The simplest call to `funnel_plot()` requires only a numerator, denominator, and group variable.

```{r}
#| label: funnelplotr-basic
#| fig-width: 10

funnel_plot(
  .data       = covid19,
  numerator   = Positive,
  denominator = Death,
  group       = `Sub-district`
)
```

::: callout-tip
**What to note:**

-   `group` defines the granularity of points (sub-district vs. district or city).
-   Default `data_type` is `"SR"` (standardised ratio).
-   The plot is automatically adjusted for overdispersion.
-   Confidence limits shown are 95% and 99.8% by default.
:::

------------------------------------------------------------------------

### Makeover 1. Switch to Proportions with Zoomed Axes

Swap numerator/denominator so we plot the **fatality rate** (deaths ÷ positives), change `data_type` to `"PR"` (proportions), and zoom the axes for readability.

```{r}
#| label: funnelplotr-makeover1
#| fig-width: 10

funnel_plot(
  .data       = covid19,
  numerator   = Death,
  denominator = Positive,
  group       = `Sub-district`,
  data_type   = "PR",        # proportions instead of standardised ratio
  xrange      = c(0, 6500),  # zoom x-axis
  yrange      = c(0, 0.05)   # zoom y-axis
)
```

::: callout-tip
`data_type = "PR"` switches the underlying statistical model to a proportion-based funnel, which is more appropriate when comparing rates (deaths per case) across units of different sizes.
:::

------------------------------------------------------------------------

### Makeover 2. Clean Labels and Title

Remove automatic outlier labels and add descriptive axis titles and a plot title.

```{r}
#| label: funnelplotr-makeover2
#| fig-width: 10

funnel_plot(
  .data       = covid19,
  numerator   = Death,
  denominator = Positive,
  group       = `Sub-district`,
  data_type   = "PR",
  xrange      = c(0, 6500),
  yrange      = c(0, 0.05),
  label       = NA,           # remove outlier labels
  title       = "Cumulative COVID-19 Fatality Rate by Cumulative Total Number of COVID-19 Positive Cases",
  x_label     = "Cumulative COVID-19 Positive Cases",
  y_label     = "Cumulative Fatality Rate"
)
```




## Building a Funnel Plot from Scratch with ggplot2

Using `FunnelPlotR` is convenient, but building the plot manually with `ggplot2` gives full control over every visual element. 

### Derive Basic Fields

Compute the **death rate** and its **standard error** for each sub-district, then filter out zero-rate sub-districts.

```{r}
#| label: derive-fields

df <- covid19 %>%
  mutate(rate    = Death / Positive,
         rate.se = sqrt((rate * (1 - rate)) / Positive)) %>%
  filter(rate > 0)

glimpse(df)
```

Compute the **weighted grand mean** rate (inverse-variance weighting):

```{r}
#| label: fit-mean

fit.mean <- weighted.mean(df$rate, 1 / df$rate.se^2)
fit.mean
```

---

### Compute 95% and 99.9% Confidence Limits

Generate a sequence of denominator values from 1 to the maximum, then compute lower/upper funnel limits at both confidence levels.

```{r}
#| label: compute-ci

number.seq   <- seq(1, max(df$Positive), 1)

number.ll95  <- fit.mean - 1.96 * sqrt((fit.mean * (1 - fit.mean)) / number.seq)
number.ul95  <- fit.mean + 1.96 * sqrt((fit.mean * (1 - fit.mean)) / number.seq)
number.ll999 <- fit.mean - 3.29 * sqrt((fit.mean * (1 - fit.mean)) / number.seq)
number.ul999 <- fit.mean + 3.29 * sqrt((fit.mean * (1 - fit.mean)) / number.seq)

dfCI <- data.frame(
  number.seq, fit.mean,
  number.ll95, number.ul95,
  number.ll999, number.ul999
)

head(dfCI)
```

::: callout-note
**Why these multipliers?**

| Limit | Multiplier | Interpretation |
|---|---|---|
| 95% CI | 1.96 | ~2 SD from mean; "common cause" variation |
| 99.9% CI | 3.29 | ~3.3 SD from mean; strong evidence of outlier |
:::

---

### Static Funnel Plot with ggplot2

Layer the data points, funnel bands, and a mean reference line.

```{r}
#| label: static-funnel-plot
#| fig-width: 10
#| fig-height: 6

p <- ggplot(df, aes(x = Positive, y = rate)) +
  
  # Data points
  geom_point(aes(label = `Sub-district`),
             alpha = 0.4) +
  
  # 95% lower limit (dashed)
  geom_line(data = dfCI,
            aes(x = number.seq, y = number.ll95),
            size = 0.4, colour = "grey40", linetype = "dashed") +
  
  # 95% upper limit (dashed)
  geom_line(data = dfCI,
            aes(x = number.seq, y = number.ul95),
            size = 0.4, colour = "grey40", linetype = "dashed") +
  
  # 99.9% lower limit (solid)
  geom_line(data = dfCI,
            aes(x = number.seq, y = number.ll999),
            size = 0.4, colour = "grey40") +
  
  # 99.9% upper limit (solid)
  geom_line(data = dfCI,
            aes(x = number.seq, y = number.ul999),
            size = 0.4, colour = "grey40") +
  
  # Grand mean reference line
  geom_hline(data = dfCI,
             aes(yintercept = fit.mean),
             size = 0.4, colour = "grey40") +
  
  # Zoom y-axis without clipping CI lines
  coord_cartesian(ylim = c(0, 0.05)) +
  
  # CI band labels
  annotate("text", x = 1,   y = -0.13, label = "95%",
           size = 3, colour = "grey40") +
  annotate("text", x = 4.5, y = -0.18, label = "99%",
           size = 3, colour = "grey40") +
  
  # Titles
  ggtitle("Cumulative Fatality Rate by Cumulative Number of COVID-19 Cases") +
  xlab("Cumulative Number of COVID-19 Cases") +
  ylab("Cumulative Fatality Rate") +
  
  theme_light() +
  theme(
    plot.title        = element_text(size = 12),
    legend.position   = c(0.91, 0.85),
    legend.title      = element_text(size = 7),
    legend.text       = element_text(size = 7),
    legend.background = element_rect(colour = "grey60", linetype = "dotted"),
    legend.key.height = unit(0.3, "cm")
  )

p
```

::: callout-tip
**Reading the funnel plot:**

- Points **within** the 95% band: variation is within expected statistical noise.
- Points **between** the 95% and 99.9% bands: moderate evidence of unusual performance.
- Points **outside** the 99.9% band: strong evidence the sub-district's rate differs from the overall average.
:::

---

### Interactive Funnel Plot with plotly

Wrap the `ggplot2` object with `ggplotly()` to enable hover tooltips showing the sub-district name, case count, and fatality rate.

```{r}
#| label: interactive-funnel-plot

fp_ggplotly <- ggplotly(p,
  tooltip = c("label", "x", "y"))

fp_ggplotly
```

::: callout-tip
Hover over any point to see the **sub-district name**, **cumulative positive cases**, and **fatality rate**. This makes it easy to identify and investigate outlying sub-districts without cluttering the static chart with labels.
:::

---

## Summary

| Method | Pros | Cons |
|---|---|---|
| `FunnelPlotR` | Fast, handles overdispersion automatically | Limited customisation |
| `ggplot2` (manual) | Full control over every visual element | More code required |
| `plotly` + `ggplot2` | Adds interactivity with minimal extra code | Requires `label` aesthetic on `geom_point` |

A funnel plot is most valuable when we need to compare many units (e.g., hospitals, districts) while accounting for the fact that **smaller units have more variability** — the narrowing funnel shape naturally encodes this statistical reality.

---

## Reference

Kam, T.S. (2023). [Funnel Plots for Fair Comparisons](https://r4va.netlify.app/chap08).



