---
title: "Hands-on Exercise 4b: Visual Statistical Analysis"
author: "Lei Yuhe"
date: "May 10, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview

Key things covered:

-   `ggstatsplot` for visual + statistical output in a single chart
-   One-sample tests, two-sample tests, ANOVA, correlation, and association tests
-   What Bayes Factors actually mean

## Setting Up

### Packages

```{r}
pacman::p_load(ggstatsplot, tidyverse)
```

[`ggstatsplot`](https://www.indrapatil.com/ggstatsplot/index.html) is built on top of [`ggplot2`](https://ggplot2.tidyverse.org/), it follows the APA standard for reporting statistics, which means the output is actually publication-ready.

### Load the Data

```{r}
exam <- read_csv("data/Exam_data.csv")
glimpse(exam)
```

The dataset has 322 students with scores in English, Maths, and Science, plus demographic info like Class, Gender, and Race.

------------------------------------------------------------------------

## One-Sample Test with `gghistostats()`

**Question**: Is the *average* English score significantly different from 60?

```{r}
set.seed(1234)

gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",       # using Bayesian inference
  test.value = 60,      # testing against this benchmark
  xlab = "English scores"
)
```

::: {.callout-note}
`type = "bayes"` runs a **Bayesian one-sample t-test** rather than the classic frequentist version.
:::

- log_e(BF₀₁) = −31.45, this is the log Bayes Factor in favour of H0. It's massively negative, meaning the evidence is overwhelmingly against H0 (that the mean = 60). Flipping it: BF₁₀ is enormous, so we have extreme evidence that the true mean is not 60.
- δ̂ posterior difference = 7.16,  the standardised effect size
- CI⁹⁵% [5.54, 8.75], 95% credible interval for that effect
- r_Cauchy = 0.71, the prior scale used for the Bayesian test

## Understanding the Bayes Factor

A **Bayes Factor (BF₁₀)** compares two hypotheses:

- **H0**: the null hypothesis (e.g., mean = 60)
- **H1**: the alternate hypothesis (e.g., mean ≠ 60)

$$BF_{10} = \frac{P(\text{data} \mid H_1)}{P(\text{data} \mid H_0)}$$
One practical way to estimate it is using the [**Schwarz/BIC criterion**](https://www.statisticshowto.com/bayesian-information-criterion/). It's a rougher approximation but much easier to compute.

### Interpreting BF₁₀ Values

| Bayes Factor | Interpretation |
|---|---|
| > 100 | Extreme evidence for H1 |
| 30 – 100 | Very strong evidence for H1 |
| 10 – 30 | Strong evidence for H1 |
| 3 – 10 | Moderate evidence for H1 |
| 1 – 3 | Anecdotal evidence for H1 |
| 1 | No evidence either way |
| 1/3 – 1 | Anecdotal evidence for H0 |
| 1/10 – 1/3 | Moderate evidence for H0 |
| < 1/10 | Strong evidence for H0 |

---

## Two-Sample Mean Test with `ggbetweenstats()`

**Question:** Do Male and Female students score differently in Maths?

### Testing for normality 
- Step 1: Test normality for each gender group
```{r}
exam %>%
  group_by(GENDER) %>%
  summarise(
    shapiro_p = shapiro.test(MATHS)$p.value,
    n = n()
  )
```

- Step 2: Visualise the distribution
```{r}
ggplot(exam, aes(sample = MATHS, color = GENDER)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~GENDER) +
  labs(title = "Q-Q Plot of Maths Scores by Gender")
```
- Step 3: Choose test based on normality result
  - If Shapiro-Wilk p > 0.05 → normality holds → use "p" (parametric, Student's t-test)
  - If Shapiro-Wilk p < 0.05 → normality violated → use "np" (non-parametric, Mann-Whitney U)


```{r}
ggbetweenstats(
  data = exam,
  x = GENDER,
  y = MATHS,
  type = "np",        # non-parametric test (Mann-Whitney U)
  messages = FALSE
)
```

`type = "np"` was used here for a **non-parametric test**, as the data violates normality assumption based on the Shapiro-Wilk's test. 
The plot shows the data distribution (violin + boxplot + points) alongside the test statistic.

---

## One-Way ANOVA with `ggbetweenstats()`

The same function handles ANOVA when there are more than 2 groups. Here I'm checking if English scores differ across racial groups.

```{r}
ggbetweenstats(
  data = exam,
  x = RACE,
  y = ENGLISH,
  type = "p",                      # parametric (standard ANOVA)
  mean.ci = TRUE,                  # show confidence intervals for means
  pairwise.comparisons = TRUE,     # run post-hoc pairwise tests
  pairwise.display = "s",          # "s" = show only significant pairs
  p.adjust.method = "fdr",         # False Discovery Rate correction
  messages = FALSE
)
```


### Options for `pairwise.display`

| Value | What it shows |
|---|---|
| `"s"` | Only significant pairwise comparisons |
| `"ns"` | Only non-significant comparisons |
| `"all"` | All pairwise comparisons |

### Which test runs under the hood?

`ggbetweenstats` is smart about picking the right test depending on `type` and group count:

| Scenario | `type = "p"` | `type = "np"` | `type = "r"` (robust) | `type = "bayes"` |
|---|---|---|---|---|
| 2 groups | Student's t | Mann-Whitney U | Yuen's t | Bayes t-test |
| 3+ groups | Fisher's ANOVA | Kruskal-Wallis | Trimmed means ANOVA | Bayes ANOVA |

Post-hoc pairwise comparisons are adjusted. FDR (`"fdr"`) is a good choice to control false discoveries when there are many comparisons.

---

## Correlation Test with `ggscatterstats()`

**Question**: Is there a significant correlation between Maths and English scores?

```{r}
ggscatterstats(
  data = exam,
  x = MATHS,
  y = ENGLISH,
  marginal = FALSE    # set TRUE to add marginal distributions on the sides
)
```

The scatterplot comes with:

- A fitted regression line
- Confidence band
- Correlation coefficient and p-value right in the subtitle

Setting `marginal = TRUE` would add density/histogram plots along the axes — useful but can be visually busy.

---

## Association Test with `ggbarstats()`

**Question**: Is there a significant association between Maths performance bands and Gender?

First, I need to bin the continuous Maths score into categories:

```{r}
exam1 <- exam %>%
  mutate(
    MATHS_bins = cut(
      MATHS,
      breaks = c(0, 60, 75, 85, 100),
      labels = c("Fail (<60)", "Pass (60-75)", "Good (75-85)", "Excellent (85+)")
    )
  )
```

Then plot:

```{r}
ggbarstats(
  data = exam1,
  x = MATHS_bins,
  y = GENDER
)
```

This runs a **chi-square test of independence** and displays:

- Stacked bar proportions by gender
- Chi-square statistic, degrees of freedom, p-value
- Cramér's V (effect size)

The `cut()` function is important here — I defined breaks at 0, 60, 75, 85, 100 to create meaningful academic performance bands.

---

## Summary

| Goal | Function | Key argument |
|---|---|---|
| Test one sample against a value | `gghistostats()` | `test.value` |
| Compare two groups | `ggbetweenstats()` | `x` = grouping (2 levels) |
| Compare 3+ groups (ANOVA) | `ggbetweenstats()` | `x` = grouping (3+ levels) |
| Test correlation | `ggscatterstats()` | `x`, `y` continuous |
| Test association (categorical) | `ggbarstats()` | `x` = binned/categorical |

## Reference
Kam, T. S. (2023) [R for Visual Analytics](https://r4va.netlify.app/chap10)
















