---
title: "Hands-on Exercise 4d: Building Funnel Plot with R"
author: "Lei Yuhe"
date: "May 3, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Funnel Plots for Fair Comparison

## Overview

Three methods for creating **funnel plots** for fair visual comparison:

1.  The `FunnelPlotR` package (quick method)
2.  `ggplot2` from scratch (manual method)
3.  `plotly` + `ggplot2` for interactivity

The dataset used is **COVID-19_DKI_Jakarta.csv**, which contains cumulative COVID-19 cases and deaths by sub-district (*kelurahan*) as at 31 July 2021.

------------------------------------------------------------------------

## Load Packages

```{r}
#| label: load-packages

pacman::p_load(tidyverse, FunnelPlotR, plotly, knitr)
```

**Packages used:**

| Package       | Purpose                                  |
|---------------|------------------------------------------|
| `tidyverse`   | Data wrangling & `ggplot2` visualisation |
| `FunnelPlotR` | High-level funnel plot creation          |
| `plotly`      | Interactive visualisation                |
| `knitr`       | Rendering static HTML tables             |

------------------------------------------------------------------------

## Import Data

```{r}
#| label: import-data

covid19 <- read_csv("data/COVID-19_DKI_Jakarta.csv") %>%
  mutate_if(is.character, as.factor)
```

::: callout-note
The dataset covers 267 sub-districts across five cities in DKI Jakarta. Key columns are `Positive` (cumulative confirmed cases) and `Death` (cumulative deaths).
:::

Preview the first few rows:

```{r}
#| label: preview-data

covid19 %>%
  head(6) %>%
  kable()
```

## FunnelPlotR Methods

### Basic Funnel Plot

The simplest call to `funnel_plot()` requires only a numerator, denominator, and group variable.

```{r}
#| label: funnelplotr-basic
#| fig-width: 10

funnel_plot(
  .data       = covid19,
  numerator   = Positive,
  denominator = Death,
  group       = `Sub-district`
)
```

::: callout-tip
**What to note:**

-   `group` defines the granularity of points (sub-district vs. district or city).
-   Default `data_type` is `"SR"` (standardised ratio).
-   The plot is automatically adjusted for overdispersion.
-   Confidence limits shown are 95% and 99.8% by default.
:::

------------------------------------------------------------------------

### Makeover 1. Switch to Proportions with Zoomed Axes

Swap numerator/denominator so we plot the **fatality rate** (deaths ÷ positives), change `data_type` to `"PR"` (proportions), and zoom the axes for readability.

```{r}
#| label: funnelplotr-makeover1
#| fig-width: 10

funnel_plot(
  .data       = covid19,
  numerator   = Death,
  denominator = Positive,
  group       = `Sub-district`,
  data_type   = "PR",        # proportions instead of standardised ratio
  xrange      = c(0, 6500),  # zoom x-axis
  yrange      = c(0, 0.05)   # zoom y-axis
)
```

::: callout-tip
`data_type = "PR"` switches the underlying statistical model to a proportion-based funnel, which is more appropriate when comparing rates (deaths per case) across units of different sizes.
:::

------------------------------------------------------------------------

### Makeover 2. Clean Labels and Title

Remove automatic outlier labels and add descriptive axis titles and a plot title.

```{r}
#| label: funnelplotr-makeover2
#| fig-width: 10

funnel_plot(
  .data       = covid19,
  numerator   = Death,
  denominator = Positive,
  group       = `Sub-district`,
  data_type   = "PR",
  xrange      = c(0, 6500),
  yrange      = c(0, 0.05),
  label       = NA,           # remove outlier labels
  title       = "Cumulative COVID-19 Fatality Rate by Cumulative Total Number of COVID-19 Positive Cases",
  x_label     = "Cumulative COVID-19 Positive Cases",
  y_label     = "Cumulative Fatality Rate"
)
```
