---
title: "Hands-on Exercise 3b: Programming Animated Statistical Graphics with R"
author: "Lei Yuhe"
date: "May 10, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Basic Concepts of Animation

When creating animations, the plot does not actually move. Instead, many individual plots are built and then stitched together as movie frames, like an old-school flip book or cartoon. Each frame is a different plot when conveying motion, built using a relevant subset of the aggregate data. The subset drives the flow of the animation when stitched back together.

Before diving in, it is important to understand some key concepts:

1. **Frame**: Each frame represents a different point in time or category. When the frame changes, the data points on the graph update to reflect the new data.

2. **Animation Attributes**: Settings that control how the animation behaves — for example, the duration of each frame, the easing function used to transition between frames, and whether to start from the current frame or the beginning.

## Getting Started

### Loading the R Packages

The following packages are used in this chapter:

- [**plotly**](https://plotly.com/r/) for plotting interactive statistical graphs
- [**gganimate**](https://gganimate.com/): a ggplot2 extension for animated statistical graphs
- [**gifski**](https://cran.r-project.org/web/packages/gifski/index.html) converts video frames to GIF animations using efficient cross-frame palettes and temporal dithering
- [**gapminder**](https://cran.r-project.org/web/packages/gapminder/index.html) provides the `country_colors` colour scheme used in the plots
- [**tidyverse**](https://www.tidyverse.org/) for data wrangling, transformation, and static graphics

```{r}
pacman::p_load(readxl, gifski, gapminder,
               plotly, gganimate, tidyverse)
```

### Importing the Data

The *Data* worksheet from the *GlobalPopulation* Excel workbook is used throughout this chapter.

```{r}
col <- c("Country", "Continent")
globalPop <- read_xls("data/GlobalPopulation.xls",
                      sheet = "Data") %>%
  mutate(across(col, as.factor)) %>%
  mutate(Year = as.integer(Year))
```

::: callout-note
**Things to learn from the code chunk above:**

- `read_xls()` from **readxl** imports the Excel worksheet.
- `mutate(across())` converts the `Country` and `Continent` character columns to factors. This replaces the deprecated `mutate_each_()` and `mutate_at()` approaches from older dplyr versions.
- `mutate(Year = as.integer(Year))` converts the Year field to integer type.
:::

---

## Animated Data Visualisation: gganimate Methods

[**gganimate**](https://gganimate.com/) extends the grammar of graphics implemented by ggplot2 to include animation. It provides a range of new grammar classes that can be added to the plot object:

| Function | Purpose |
|---|---|
| `transition_*()` | Defines how data spreads across time |
| `view_*()` | Defines how positional scales change during animation |
| `shadow_*()` | Defines how data from other time points is displayed |
| `enter_*()/exit_*()` | Defines how new/old data appears and disappears |
| `ease_aes()` | Defines how aesthetics are eased during transitions |

### Building a Static Population Bubble Plot

The basic ggplot2 functions are used to create a static bubble plot first. This is the foundation before animating.

```{r}
#| fig-width: 7
#| fig-height: 4

ggplot(globalPop, aes(x = Old, y = Young,
                      size = Population,
                      colour = Country)) +
  geom_point(alpha = 0.7,
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}',
       x = '% Aged',
       y = '% Young')
```


### Building the Animated Bubble Plot

Two gganimate functions are added to bring the plot to life:

- [`transition_time()`](https://gganimate.com/reference/transition_time.html): creates transitions through distinct states in time (i.e. Year)
- `ease_aes()`: controls easing of aesthetics during transitions. The default is `linear`, other options include quadratic, cubic, sine, elastic, bounce, and more.


```{r}
#| fig-width: 7
#| fig-height: 4

ggplot(globalPop, aes(x = Old, y = Young,
                      size = Population,
                      colour = Country)) +
  geom_point(alpha = 0.7,
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}',
       x = '% Aged',
       y = '% Young') +
  transition_time(Year) +
  ease_aes('linear')
```


---

## Animated Data Visualisation: plotly

In **Plotly R**, both `ggplotly()` and `plot_ly()` support key frame animations through the `frame` argument/aesthetic. They also support an `ids` argument/aesthetic to ensure smooth transitions between objects with the same id (facilitating *object constancy*).

### Building an Animated Bubble Plot: `ggplotly()` Method

The animated bubble plot produced with `ggplotly()` includes a play/pause button and a slider component for controlling the animation.

```{r}
#| fig-width: 7
#| fig-height: 4


gg <- ggplot(globalPop,
       aes(x = Old,
           y = Young,
           size = Population,
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged',
       y = '% Young') +
  theme(legend.position = 'none')

ggplotly(gg)
```

::: callout-note
**Things to learn from the code chunk above:**

- A static bubble plot is built with ggplot2 and saved as the R object `gg`.
- `ggplotly()` converts the R graphic object into an animated SVG object.
- `theme(legend.position = 'none')` is required to suppress the legend. 
- Using `show.legend = FALSE` inside `geom_point()` does not remove it in plotly.
:::

### Building an Animated Bubble Plot: `plot_ly()` Method

The `plot_ly()` method provides a more direct plotly interface, colouring bubbles by Continent and using `hoverinfo` to show country names on hover.

```{r}
bp <- globalPop %>%
  plot_ly(x = ~Old,
          y = ~Young,
          size = ~Population,
          color = ~Continent,
          sizes = c(2, 100),
          frame = ~Year,
          text = ~Country,
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers'
          ) %>%
  layout(showlegend = FALSE)

bp
```

---


## Reference

Kam, T. S. (2023) [R for Visual Analytics](https://r4va.netlify.app/chap04)






