---
title: "Hands-on Exercise 2"
author: "Lei Yuhe"
date: "April 19, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
format: html
---

## Beyond ggplot2 Fundamentals
Beside tidyverse, four R packages will be used. Here is a table for the packages, their purpose and primary functions: 

| Package | Purpose | Primary Functions |
|---|---|---|
| `ggrepel` | Repel overlapping text and label geoms | `geom_text_repel()`, `geom_label_repel()` |
| `patchwork` | Compose multiple ggplots into a single figure | `+`, `/`, `\|`, `plot_layout()`, `plot_annotation()` |
| `ggthemes` | Extra themes inspired by news and science outlets | `theme_economist()`, `theme_wsj()`, `theme_tufte()` |
| `hrbrthemes` | Typography-centric, publication-ready themes | `theme_ipsum()`, `theme_ipsum_rc()`, `theme_modern_rc()` |
| `tidyverse` | Core data wrangling and ggplot2 visualisation | `ggplot()`, `dplyr`, `tidyr`, `readr` |

Check the installation: 
```{r}
pacman::p_load(ggrepel, patchwork, 
               ggthemes, hrbrthemes,
               tidyverse) 
```

Import the data: 
```{r}
exam_data <- read_csv("data/Exam_data.csv")
```


## Beyond ggplot2 Annotation: ggrepel

One of the challenge in plotting statistical graph is annotation, especially with large number of data points.

::: {.panel-tabset}

## The plot

```{r, echo=FALSE}
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              size = 0.5) +
  geom_label(aes(label = ID),
             hjust = .5,
             vjust = -.5) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## The code

```r
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              size = 0.5) +
  geom_label(aes(label = ID),
             hjust = .5,
             vjust = -.5) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

:::

## Working with ggrepel

::: {.panel-tabset}

## The plot

```{r, echo=FALSE}
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              size = 0.5) +
  geom_label_repel(aes(label = ID),
                   fontface = "bold") +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## The code

```r
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              size = 0.5) +
  geom_label_repel(aes(label = ID),
                   fontface = "bold") +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

:::


## Beyond ggplots Themes: ggthemes & hrbrthemes

ggplot2 comes with eight [built-in themes](https://ggplot2.tidyverse.org/reference/ggtheme.html), they are: `theme_gray()`, `theme_bw()`, `theme_classic()`, `theme_dark()`, `theme_light()`, `theme_linedraw()`, `theme_minimal()`, and `theme_void()`.

::: {.panel-tabset}

## The plot

```{r, echo=FALSE}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores")
```

## The code

```r
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores")
```

:::

Refer to this [link](https://ggplot2.tidyverse.org/reference/ggtheme.html) to learn more about ggplot2 `Themes`

## Working with ggtheme

[ggthemes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) provides ['ggplot2' themes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) that replicate the look of plots by Edward Tufte, Stephen Few, [Fivethirtyeight](https://fivethirtyeight.com/), [The Economist](https://www.economist.com/), 'Stata', 'Excel', and [The Wall Street Journal](https://www.wsj.com/), among others.

In the example below, *The Economist* theme is used.

::: {.panel-tabset}

## The plot

```{r, echo=FALSE}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```

## The code

```r
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```

:::


## Working with hrbrthemes

::: {.panel-tabset}

## The plot

```{r, echo=FALSE}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

## The code

```r
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

:::

::: {.callout-tip}
## What can we learn from the code chunk above?

- `axis_title_size` argument is used to increase the font size of the axis title to 18,
- `base_size` argument is used to increase the default axis label to 15, and
- `grid` argument is used to remove the x-axis grid lines.
:::


## Beyond Single Graph

Plotting below 3 single graphs for composition. 

::: {.panel-tabset}

## The plot

```{r, echo=FALSE}
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")

p1
```

## The code

```r
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")
```
:::

::: {.panel-tabset}
## The plot

```{r, echo=FALSE}
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")

p2
```

## The code

```r
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")
```
:::

::: {.panel-tabset}
## The plot

```{r, echo=FALSE}
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")

p3
```

## The code

```r
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

## Creating Composite Graphics: pathwork methods

::: {.panel-tabset}
## The plot

```{r, echo=FALSE}
(p1 / p2) | p3
```

## The code

```r
(p1 / p2) | p3
```
:::




