---
title: "5c. Visual Multivariate Analysis with Heatmap"
author: "Lei Yuhe"
date: "May 14, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---
## Overview

Heatmaps encode data values as **colour variations** across a grid, making them excellent for exploring multivariate datasets. When applied to a tabular format, rows typically represent observations (e.g., countries) and columns represent variables (e.g., happiness indicators).

**What heatmaps are good for:**

- Showing variance across multiple variables at a glance
- Revealing clusters and patterns across rows and columns
- Detecting correlations and similarities between variables
- Comparing many observations simultaneously

This chapter covers both **static** and **interactive** heatmaps using R.

---

## Installing and Launching R Packages

Four packages are needed for this chapter:

```{r}
pacman::p_load(seriation, dendextend, heatmaply, tidyverse)
```

| Package | Role |
|---------|------|
| `tidyverse` | Data import and wrangling |
| `heatmaply` | Interactive heatmaps |
| `dendextend` | Dendrogram analysis and optimal cluster selection |
| `seriation` | Optimal row/column ordering algorithms |

---

## Importing and Preparing the Data Set

The dataset used is the **World Happiness Report 2018** (`WHData-2018.csv`).

### Importing the data set

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

`read_csv()` from `readr` imports the file as a tibble data frame stored in `wh`.

### Preparing the data

Row names are set to country names (instead of row numbers) to make the heatmap axis labels meaningful:

```{r}
row.names(wh) <- wh$Country
```

### Transforming the data frame into a matrix

Heatmap functions require a **matrix** object, not a data frame. We select the relevant numeric columns and convert:

```{r}
wh1 <- dplyr::select(wh, c(3, 7:12))   # select numeric columns of interest
wh_matrix <- data.matrix(wh)            # convert to matrix
```

> **Key concept:** `data.matrix()` coerces all columns to numeric, producing an R matrix suitable for heatmap functions.

---

## Static Heatmap

Several R packages can draw static heatmaps:

| Function / Package | Description |
|---|---|
| `heatmap()` — base R `stats` | Simple heatmap with optional dendrograms |
| `heatmap.2()` — `gplots` | Enhanced version of base heatmap |
| `pheatmap()` — `pheatmap` | "Pretty Heatmap" with more layout control |
| `ComplexHeatmap` — Bioconductor | Complex, annotated heatmaps (genomics-oriented) |
| `superheat` | Highly customisable; handles big and multi-type data |

This chapter focuses on **`heatmap()`** from base R stats.

### `heatmap()` of R Stats

**Without clustering (plain heatmap):**

```{r}
wh_heatmap <- heatmap(wh_matrix,
                      Rowv = NA, Colv = NA)
```

- `Rowv = NA` and `Colv = NA` suppress row and column dendrograms.
- The result is a simple colour grid in the original data order.

**With clustering (default behaviour):**

```{r}
wh_heatmap <- heatmap(wh_matrix)
```

- By default, `heatmap()` reorders rows and columns using **hierarchical clustering** (distance-based).
- Dendrograms are drawn on the sides to reflect the clustering structure.
- The reordering groups similar countries and variables together.

**With column normalisation (recommended):**

```{r}
wh_heatmap <- heatmap(wh_matrix,
                      scale  = "column",
                      cexRow = 0.6,
                      cexCol = 0.8,
                      margins = c(10, 4))
```

> **Why normalise?** Variables like "Happiness Score" have much larger raw values than others (e.g., "Generosity"). Without scaling, the colour variation is dominated by high-range variables, making all others look identical. Column scaling (subtract mean, divide by SD) brings variables onto a comparable scale.

**Key arguments:**

| Argument | Purpose |
|---|---|
| `scale = "column"` | Normalise each column (z-score) |
| `cexRow` / `cexCol` | Font size for row / column labels |
| `margins = c(col, row)` | Extra space for long axis labels |

---

## Creating Interactive Heatmap

[`heatmaply`](http://talgalili.github.io/heatmaply/index.html) builds interactive cluster heatmaps exportable as standalone HTML. It wraps `plotly` under the hood, enabling zoom, hover tooltips, and pan.

### Working with `heatmaply`

Basic usage:

```{r}
#| fig-width: 10
#| fig-height: 10
heatmaply(wh_matrix[, -c(1, 2, 4, 5)])
```

**Differences from base `heatmap()`:**

- The horizontal dendrogram appears on the **left** side (not top).
- Row text labels are on the **right** side.
- Long x-axis labels are auto-rotated 135°.

---

### Data Transformation

Because multivariate datasets often mix variables with different units and ranges, **data transformation before clustering is essential**. `heatmaply` supports three methods:

#### Scaling (Z-score)

**When to use:** Variables are assumed to come from (approximately) normal distributions.

**How it works:** Subtract the column mean and divide by the standard deviation → values represent distance from the mean in SD units.

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)],
          scale = "column")
```

#### Normalising (Min-Max)

**When to use:** Variables come from different, possibly non-normal distributions.

**How it works:** Subtract the minimum, divide by the maximum → rescales each variable to [0, 1]. Preserves the shape of each variable's distribution.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]))
```

#### Percentising (Empirical CDF)

**When to use:** You want each value to have a clear percentile interpretation.

**How it works:** Replaces each value with its empirical percentile rank. A value of 0.75 means 75% of observations are at or below it.

```{r}
heatmaply(percentize(wh_matrix[, -c(1, 2, 4, 5)]))
```

> **Summary of transformation choices:**
>
> | Method | Input assumption | Output range | Interpretation |
> |---|---|---|---|
> | Scale | Normal distribution | ~(-3, 3) | SD from mean |
> | Normalize | Any distribution | [0, 1] | Relative magnitude |
> | Percentize | Any distribution | [0, 1] | Percentile rank |

---

### Clustering Algorithm

`heatmaply` supports hierarchical clustering with configurable distance and linkage methods.

**Key arguments:**

| Argument | Default | Options |
|---|---|---|
| `distfun` | `dist` (Euclidean) | `"pearson"`, `"spearman"`, `"kendall"` for correlation-based |
| `dist_method` | `"euclidean"` | `"maximum"`, `"manhattan"`, `"canberra"`, `"binary"`, `"minkowski"` |
| `hclustfun` | `hclust` | — |
| `hclust_method` | `"complete"` | `"ward.D"`, `"ward.D2"`, `"single"`, `"average"`, `"mcquitty"`, `"median"`, `"centroid"` |

Clustering can be calibrated **manually** or **statistically**.

---

### Manual Approach

Specify distance and linkage methods directly:

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method  = "euclidean",
          hclust_method = "ward.D")
```

Ward.D minimises total within-cluster variance, tending to produce compact, roughly equal-sized clusters.

---

### Statistical Approach

Use `dendextend` functions to **objectively select** the best method and number of clusters.

**Step 1 — Find the best clustering method using `dend_expend()`:**

```{r}
wh_d <- dist(normalize(wh_matrix[, -c(1, 2, 4, 5)]), method = "euclidean")
dend_expend(wh_d)[[3]]
```

This compares multiple linkage methods by their optimisation score. The method with the **highest `optim` value** is recommended. For the WHR 2018 data, `"average"` (UPGMA) gives the best result.

**Step 2 — Find the optimal number of clusters using `find_k()`:**

```{r}
wh_clust <- hclust(wh_d, method = "average")
num_k <- find_k(wh_clust)
plot(num_k)
```

The plot suggests **k = 3** as the optimal number of clusters.

**Step 3 — Apply the statistical findings:**

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method   = "euclidean",
          hclust_method = "average",
          k_row         = 3)
```

`k_row = 3` highlights the three clusters with coloured row side-bars.

---

### Seriation

Hierarchical clustering determines which items are *near* each other in the tree, but leaves the **branch flip order** ambiguous. Seriation finds the optimal leaf ordering to minimise visual jumps between adjacent rows/columns.

> **Analogy:** Think of it as solving a restricted Travelling Salesman Problem — find the path through all leaves that minimises total distance, subject to the dendrogram structure.

`heatmaply` uses the `seriation` package and offers four `seriate` options:

| Option | Algorithm | Notes |
|---|---|---|
| `"OLO"` | Optimal Leaf Ordering | Default; exact optimum in O(n⁴) |
| `"GW"` | Gruvaeus & Wainer | Faster heuristic; near-optimal |
| `"mean"` | Mean-based | Matches default behaviour of `gplots::heatmap.2` |
| `"none"` | No rotation | Uses raw dendrogram without reordering |

```{r}
# Optimal Leaf Ordering (default)
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]), seriate = "OLO")

# Gruvaeus-Wainer heuristic
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]), seriate = "GW")

# Mean-based (gplots-compatible)
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]), seriate = "mean")

# No seriation
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]), seriate = "none")
```

---

### Working with Colour Palettes

The default palette in `heatmaply` is **viridis** (perceptually uniform, colourblind-friendly). Alternative palettes from `RColorBrewer` can be specified via the `colors` argument:

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none",
          colors  = Blues)
```

Choose palettes that reflect the data's nature:

- **Sequential** (e.g., Blues, Greens): for data with a natural order from low to high.
- **Diverging** (e.g., RdBu): for data with a meaningful midpoint (e.g., z-scores).
- **Qualitative**: not appropriate for continuous heatmap cells.

---

### The Finishing Touch

`heatmaply` provides extensive arguments for publication-quality output:

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          Colv         = NA,
          seriate      = "none",
          colors       = Blues,
          k_row        = 5,
          margins      = c(NA, 200, 60, NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main = "World Happiness Score and Variables by Country, 2018\nData Transformation using Normalise Method",
          xlab = "World Happiness Indicators",
          ylab = "World Countries")
```

**Key finishing arguments:**

| Argument | Purpose |
|---|---|
| `k_row` | Number of row cluster groups to highlight |
| `margins` | Margins in order: left, right, top, bottom |
| `fontsize_row` / `fontsize_col` | Label font sizes (use small values for many rows) |
| `main` | Plot title (`\n` for line breaks) |
| `xlab` / `ylab` | Axis titles |
| `Colv = NA` | Suppress column dendrogram |

---

## Key Takeaways

1. **Always transform your data** before clustering in a heatmap. The right method depends on your data's distribution — use scale for normal, normalize for bounded, percentize for rank-based interpretation.

2. **Use statistical methods** (`dend_expend()` + `find_k()`) to objectively choose clustering method and number of clusters rather than guessing.

3. **Seriation improves readability** — OLO or GW ordering makes patterns clearer than default dendrogram ordering.

4. **`heatmaply` > base `heatmap()`** for most analytical work: it's interactive, supports transformations natively, integrates seriation, and allows fine-grained aesthetic control.

5. **Colour palette choice matters** — viridis is the safe default; Blues or other sequential palettes work for normalised (0–1) data; diverging palettes suit z-scores.

---

## Packages Reference

```{r}
# Load all packages for this chapter
pacman::p_load(seriation, dendextend, heatmaply, tidyverse)
```

| Package | Key Functions Used |
|---|---|
| `tidyverse` / `readr` | `read_csv()` |
| `dplyr` | `select()` |
| `stats` (base R) | `heatmap()`, `dist()`, `hclust()`, `data.matrix()` |
| `heatmaply` | `heatmaply()`, `normalize()`, `percentize()` |
| `dendextend` | `dend_expend()`, `find_k()` |
| `seriation` | Used internally by `heatmaply` for OLO/GW seriation |