---
title: "5a. Creating Ternary Plot with R"
author: "Lei Yuhe"
date: "May 14, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Packages

Two main packages are used:

- [`ggtern`](http://www.ggtern.com/): extends `ggplot2` to support ternary coordinate systems
- [`plotly`](https://plotly.com/r/): creates interactive web-based charts, its `scatterternary` trace type renders interactive ternary diagrams

Supporting tidyverse packages (`readr`, `dplyr`, `tidyr`) handle data import and wrangling.

```{r}
#| label: load-packages

pacman::p_load(plotly, ggtern, tidyverse)
```

> **Note:** `ggtern` masks several `ggplot2` functions (e.g. `annotate()`). Load it after tidyverse to avoid conflicts, and reload `ggplot2` functions with `ggplot2::` prefix if needed later.

## Data Preparation

### The Dataset

**Singapore Residents by Planning Area/Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020**
Source: [SingStat](https://www.singstat.gov.sg/find-data/explore-data-themes/population/geographic-distribution/latest-news-data)

File: `data/respopagesextod2011to2020`

| Column | Description |
|--------|-------------|
| `PA` | Planning Area |
| `SZ` | Subzone |
| `AG` | Age Group |
| `Sex` | Sex |
| `TOD` | Type of Dwelling |
| `Population` | Resident count |
| `Year` | Reference year (2011 - 2020) |

### Import Data

```{r}
#| label: import-data

pop_data <- read_csv("data/respopagesextod2011to2020.csv")

glimpse(pop_data)
```


### Derive Age-Group Aggregates

Three composite measures are derived by summing across age-group columns after pivoting wide:

| Measure | Age bands included |
|---------|--------------------|
| `YOUNG` | 0–4, 5–9, 10–14, 15–19, 20–24 (columns 4–8) |
| `ACTIVE` | 25–29, … , 55–59, 60–64 (columns 9–16) |
| `OLD` | 65–69, 70–74, 75–79, 80–84, 85+ (columns 17–21) |


```{r}
#| label: wrangle-data

#Deriving the young, economy active and old measures
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)

glimpse(agpop_mutated)
```

> **Add-on tip:** To make the workflow more robust, replace column-index slicing (`.[4:8]`) with explicit column name ranges using `dplyr::across()` once you have verified the column order.

---




