---
title: "5a. Creating Ternary Plot with R"
author: "Lei Yuhe"
date: "May 14, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Packages

Two main packages are used:

- [`ggtern`](http://www.ggtern.com/): extends `ggplot2` to support ternary coordinate systems
- [`plotly`](https://plotly.com/r/): creates interactive web-based charts, its `scatterternary` trace type renders interactive ternary diagrams

Supporting tidyverse packages (`readr`, `dplyr`, `tidyr`) handle data import and wrangling.

```{r}
#| label: load-packages

pacman::p_load(plotly, ggtern, tidyverse)
```

> **Note:** `ggtern` masks several `ggplot2` functions (e.g. `annotate()`). Load it after tidyverse to avoid conflicts, and reload `ggplot2` functions with `ggplot2::` prefix if needed later.

## Data Preparation

### The Dataset

**Singapore Residents by Planning Area/Subzone, Age Group, Sex and Type of Dwelling, June 2000-2018**
Source: [SingStat](https://www.singstat.gov.sg/find-data/explore-data-themes/population/geographic-distribution/latest-news-data)

File: `data/respopagsex2000to2018_tidy.csv`

| Column | Description |
|--------|-------------|
| `PA` | Planning Area |
| `SZ` | Subzone |
| `AG` | Age Group |
| `Sex` | Sex |
| `TOD` | Type of Dwelling |
| `Population` | Resident count |
| `Year` | Reference year (2011 - 2020) |

### Import Data

```{r}
#| label: import-data

pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv")

glimpse(pop_data)
```


### Derive Age-Group Aggregates

Three composite measures are derived by summing across age-group columns after pivoting wide:

| Measure | Age bands included |
|---------|--------------------|
| `YOUNG` | 0–4, 5–9, 10–14, 15–19, 20–24 (columns 4–8) |
| `ACTIVE` | 25–29, … , 55–59, 60–64 (columns 9–16) |
| `OLD` | 65–69, 70–74, 75–79, 80–84, 85+ (columns 17–21) |


```{r}
#| label: wrangle-data

#Deriving the young, economy active and old measures
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)

glimpse(agpop_mutated)
```

> **Add-on tip:** To make the workflow more robust, replace column-index slicing (`.[4:8]`) with explicit column name ranges using `dplyr::across()` once you have verified the column order.

---

## Plotting Ternary Diagrams

### Static Ternary Plot - Minimal

`ggtern()` follows the same grammar as `ggplot2` but uses a ternary coordinate system with axes `x`, `y`, and `z`.

```{r}
#| label: static-basic
#| fig-cap: "Basic static ternary plot — population structure by subzone, 2018"

ggtern(data = agpop_mutated,
       aes(x = YOUNG, y = ACTIVE, z = OLD)) +
  geom_point()
```

### Static Ternary Plot - Styled

Adding a title and the `theme_rgbw()` theme colours the axes Red / Green / Blue / White for easier reading.

```{r}
#| label: static-styled
#| fig-cap: "Styled static ternary plot with theme_rgbw()"

ggtern(data = agpop_mutated,
       aes(x = YOUNG, y = ACTIVE, z = OLD)) +
  geom_point(alpha = 0.6, size = 2, colour = "steelblue") +
  labs(title    = "Population Structure by Subzone, 2018",
       subtitle = "Singapore Residents — Young / Active / Old composition",
       x = "Young", y = "Active", z = "Old") +
  theme_rgbw()
```


#### Add-on: Colour-encode a grouping variable {.unnumbered}

If planning areas (`PA`) are available after the spread, you can map colour to a categorical variable to reveal spatial clusters:

```{r}
#| label: static-coloured
#| eval: false  # set to true if PA column is present

ggtern(data = agpop_mutated,
       aes(x = YOUNG, y = ACTIVE, z = OLD, colour = PA)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "Population Structure by Planning Area, 2018") +
  theme_rgbw() +
  theme(legend.position = "right")
```

### Interactive Ternary Plot with Plotly

`plot_ly()` with `type = "scatterternary"` renders an interactive version where you can hover to inspect individual subzones.

```{r}
#| label: interactive-plotly
#| fig-cap: "Interactive ternary plot — hover to see subzone details"

# Helper: annotation label
label <- function(txt) {
  list(
    text = txt,
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper",
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# Helper: axis formatting
axis <- function(txt) {
  list(title = txt, tickformat = ".0%", tickfont = list(size = 10))
}

ternaryAxes <- list(
  aaxis = axis("Young"),
  baxis = axis("Active"),
  caxis = axis("Old")
)

plot_ly(
  agpop_mutated,
  a     = ~YOUNG,
  b     = ~ACTIVE,
  c     = ~OLD,
  color = I("steelblue"),
  type  = "scatterternary",
  mode  = "markers",
  text  = ~paste("Subzone:", SZ, "<br>Total:", TOTAL),  # tooltip add-on
  hoverinfo = "text"
) %>%
  layout(
    title       = "Population Structure by Subzone, Singapore 2018",
    annotations = label("Ternary Markers"),
    ternary     = ternaryAxes
  )
```


#### Add-on: Colour by population size {.unnumbered}

Map marker colour to `TOTAL` population for a richer visual encoding:

```{r}
#| label: interactive-coloured
#| fig-cap: "Interactive ternary plot — colour encodes total population"

plot_ly(
  agpop_mutated,
  a          = ~YOUNG,
  b          = ~ACTIVE,
  c          = ~OLD,
  color      = ~TOTAL,
  colors     = "YlOrRd",
  size       = ~TOTAL,
  sizes      = c(5, 30),
  type       = "scatterternary",
  mode       = "markers",
  text       = ~paste0("Subzone: ", SZ,
                       "<br>Young: ",  scales::percent(YOUNG  / TOTAL, 0.1),
                       "<br>Active: ", scales::percent(ACTIVE / TOTAL, 0.1),
                       "<br>Old: ",    scales::percent(OLD    / TOTAL, 0.1),
                       "<br>Total: ",  formatC(TOTAL, format = "d", big.mark = ",")),
  hoverinfo  = "text"
) %>%
  layout(
    title   = "Population Structure by Subzone, Singapore 2018",
    ternary = ternaryAxes,
    coloraxis = list(colorscale = "YlOrRd",
                     colorbar   = list(title = "Total Population"))
  )
```

---

## Summary

| Approach | Package | Interactivity | Best for |
|----------|---------|---------------|----------|
| `ggtern()` + `geom_point()` | `ggtern` | Static | Reports, print |
| `ggtern()` + `theme_rgbw()` | `ggtern` | Static | Styled static output |
| `plot_ly()` scatterternary | `plotly` | Interactive | Web / HTML output |

**Key takeaways:**

- Ternary plots are ideal when three proportions must sum to 1 (or 100%).
- `ggtern` extends the familiar `ggplot2` API — most `geom_*` and `theme_*` functions still work.
- `plotly`'s `scatterternary` type adds hover tooltips and zoom for exploratory analysis.
- Encoding an additional variable (colour, size) can reveal structure that a plain ternary plot obscures.








