---
title: "5d. Visual Multivariate Analysis with Parallel Coordinates Plot"
author: "Lei Yuhe"
date: "May 14, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---
## Overview

A **parallel coordinates plot (PCP)** is a multivariate visualisation technique where each variable is represented as a vertical axis, and each observation is drawn as a polyline connecting its values across all axes. It is specially designed for comparing multiple numerical variables simultaneously and surfacing relationships between them.

**Key characteristics:**

- Invented by **Alfred Inselberg** in the 1970s for high-dimensional data visualisation.
- Better suited for **analytical exploration** than for communication to general audiences. As Stephen Few (2006) noted, its strength lies in revealing meaningful multivariate patterns and comparisons when used interactively — not in presenting conclusions to a board or the public.
- Practical use case: characterising clusters found during customer segmentation, or comparing indicators like those in the World Happiness Index.

**Two approaches towards multivariate analysis using parallel coordinate plots:**

| Approach | Package | Function |
|---|---|---|
| Static PCP | `GGally` | `ggparcoord()` |
| Interactive PCP | `parallelPlot` | `parallelPlot()` |

---

## Installing and Launching R Packages

```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```

| Package | Role |
|---|---|
| `GGally` | Extends ggplot2; provides `ggparcoord()` for static PCPs |
| `parallelPlot` | Builds interactive PCPs via htmlwidgets and D3.js |
| `tidyverse` | Data import and wrangling (`readr`, `dplyr`) |

---

## Data Preparation

The same **World Happiness Report 2018** dataset from Chapter 14 is used (`WHData-2018.csv`).

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

The tibble `wh` contains columns for country name, region, happiness score, and six contributing variables (columns 7–12):

- GDP per capita
- Social support
- Healthy life expectancy
- Freedom to make life choices
- Generosity
- Perceptions of corruption

No additional matrix transformation is needed here, `ggparcoord()` and `parallelPlot()` work directly with data frames.

---

## Plotting Static Parallel Coordinates Plot

Static PCPs are built using `ggparcoord()` from the **GGally** package. Because `GGally` extends `ggplot2`, standard ggplot2 functions like `facet_wrap()` and `theme()` can be combined with it.

### Plotting a Simple Parallel Coordinates

```{r}
ggparcoord(data    = wh,
           columns = c(7:12))
```

- `data`: the data frame.
- `columns`: integer vector of column indices to use as parallel axes.

> This minimal plot is hard to read — all lines share the same colour, variables are on different scales, and there is no grouping. The following sections progressively improve it.

---

### Plotting a Parallel Coordinates with Boxplot

```{r}
#| fig-width: 14
#| fig-height: 7
ggparcoord(data        = wh,
           columns     = c(7:12),
           groupColumn = 2,
           scale       = "uniminmax",
           alphaLines  = 0.2,
           boxplot     = TRUE,
           title       = "Parallel Coordinates Plot of World Happiness Variables")
```

**Arguments explained:**

| Argument | Value | Effect |
|---|---|---|
| `groupColumn` | `2` (Region) | Colours each polyline by the observation's region |
| `scale` | `"uniminmax"` | Scales each variable so its min = 0 and max = 1; enables fair comparison across axes |
| `alphaLines` | `0.2` | Sets line transparency (0 = invisible, 1 = fully opaque); reduces overplotting clutter |
| `boxplot` | `TRUE` | Overlays a boxplot on each axis to show the variable's distribution |
| `title` | string | Adds a plot title |

> **Why `uniminmax`?** Without scaling, axes with larger raw ranges (e.g., GDP) dominate visually. `uniminmax` brings all variables to the same [0, 1] range without assuming normality (unlike z-score scaling).

**Other `scale` options available in `ggparcoord()`:**

| Value | Description |
|---|---|
| `"std"` | Subtract mean, divide by SD (z-score) |
| `"robust"` | Subtract median, divide by MAD |
| `"uniminmax"` | Min = 0, Max = 1 per variable |
| `"globalminmax"` | Scales using global min/max across all selected variables |
| `"center"` | Centers each variable at a chosen `scaleSummary` statistic |
| `"centerObs"` | Centers on a specific observation |

---

### Parallel Coordinates with Facet

Since `ggparcoord()` returns a `ggplot` object, ggplot2 functions can be appended with `+`.

`facet_wrap()` splits the plot into small multiples — one panel per region:

```{r}
#| fig-width: 14
#| fig-height: 7
ggparcoord(data        = wh,
           columns     = c(7:12),
           groupColumn = 2,
           scale       = "uniminmax",
           alphaLines  = 0.2,
           boxplot     = TRUE,
           title       = "Multiple Parallel Coordinates Plots of World Happiness Variables by Region") +
  facet_wrap(~ Region)
```

- Each of the 10 panels shows the parallel coordinates for one geographic region.
- This makes within-region patterns much clearer than a single overcrowded plot.

> **Aesthetic issue:** With many variables, the x-axis labels overlap. The next two sections fix this.

---

### Rotating x-axis Text Label

Use `theme()` with `axis.text.x` to rotate labels:

```{r}
#| fig-width: 14
#| fig-height: 7
ggparcoord(data        = wh,
           columns     = c(7:12),
           groupColumn = 2,
           scale       = "uniminmax",
           alphaLines  = 0.2,
           boxplot     = TRUE,
           title       = "Multiple Parallel Coordinates Plots of World Happiness Variables by Region") +
  facet_wrap(~ Region) +
  theme(axis.text.x = element_text(angle = 30))
```

- `angle = 30`: rotates x-axis labels 30° from horizontal.

> **Problem:** Rotating without adjusting position causes labels to overlap with the plot area.

---

### Adjusting the Rotated x-axis Text Label

Fix overlapping by adding `hjust = 1` to shift the text anchor to the right end of the label:

```{r}
#| fig-width: 14
#| fig-height: 7
ggparcoord(data        = wh,
           columns     = c(7:12),
           groupColumn = 2,
           scale       = "uniminmax",
           alphaLines  = 0.2,
           boxplot     = TRUE,
           title       = "Multiple Parallel Coordinates Plots of World Happiness Variables by Region") +
  facet_wrap(~ Region) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

- `hjust = 1`: right-justifies the text, aligning the label's right edge with the axis tick, preventing overlap with the plot area.

> **Rule of thumb:** Whenever rotating axis text with `angle > 0`, always pair it with `hjust = 1` (and sometimes `vjust = 1`) to keep labels properly anchored.

---

## Plotting Interactive Parallel Coordinates Plot: `parallelPlot` Methods

The **`parallelPlot`** package renders interactive parallel coordinates using **htmlwidgets** and **D3.js**. Key interactive features include:

- **Clicking on an axis** recolours all lines by that variable's values (intensity-based colour encoding).
- **Brushing** on any axis filters the visible lines to observations within the selected range.

### Data Preparation for `parallelPlot`

Before plotting, select only the columns of interest (all must be numeric):

```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))
```

---

### The Basic Plot

```{r}
#| fig-width: 10
#| fig-height: 10
parallelPlot(wh,
             width  = 320,
             height = 250)
```

- `width` / `height`: pixel dimensions of the widget.
- Default colour: monotonous blue for all lines.
- **Issue:** Long axis labels may overlap — fixed in the next step.

---

### Rotate Axis Label

```{r}
parallelPlot(wh,
             rotateTitle = TRUE)
```

- `rotateTitle = TRUE`: rotates all axis labels 90° to prevent overlap.

> **Interactive tip:** After rotating, clicking on any axis variable (e.g., "Happiness score") changes the colour scheme from uniform blue to a graduated blue intensity — lines are coloured by that variable's value, immediately revealing correlations.

---

### Changing the Colour Scheme

Replace the default blue with any named D3/ColorBrewer continuous palette:

```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle  = TRUE)
```

- `continuousCS`: sets the continuous colour scale used when an axis is clicked.
- `"YlOrRd"` (Yellow-Orange-Red) is a sequential palette that clearly encodes magnitude — high values appear red, low values appear yellow.

---

### Parallel Coordinates Plot with Histogram

Histograms can be displayed along each axis to show marginal distributions:

```{r}
histoVisibility <- rep(TRUE, ncol(wh))

parallelPlot(wh,
             rotateTitle     = TRUE,
             histoVisibility = histoVisibility)
```

- `histoVisibility`: a logical vector with one element per column. `TRUE` shows a histogram for that axis; `FALSE` hides it.
- `rep(TRUE, ncol(wh))` enables histograms for all columns.
- This gives an at-a-glance view of each variable's distribution alongside the parallel lines.

---

## Comparing `ggparcoord()` vs `parallelPlot()`

| Feature | `ggparcoord()` (GGally) | `parallelPlot()` (parallelPlot) |
|---|---|---|
| Output type | Static ggplot2 object | Interactive HTML widget |
| Scaling | Built-in via `scale` argument | Manual pre-processing required |
| Grouping / colour | Via `groupColumn` | Via axis click (interactive) |
| Boxplots on axes | Yes (`boxplot = TRUE`) | No |
| Faceting | Yes (via `facet_wrap`) | No |
| Histograms on axes | No | Yes (`histoVisibility`) |
| Brushing / filtering | No | Yes (D3-powered) |
| Best for | Publication figures, faceted comparisons | Exploratory analysis, interactive dashboards |

---

## Key Takeaways

1. **Parallel coordinates are for exploration, not communication.** They excel at revealing multivariate patterns when used interactively but are not suitable for general audiences or executive presentations.

2. **Always scale your variables.** Raw values on different ranges make PCPs unreadable. `uniminmax` in `ggparcoord()` is the most intuitive choice; it preserves the shape of each variable without assuming normality.

3. **Use `groupColumn` to add colour.** Colouring lines by a categorical variable (e.g., region) is the fastest way to spot group-level patterns in a static PCP.

4. **`facet_wrap()` works with `ggparcoord()`** because it returns a ggplot2 object — leverage the full ggplot2 ecosystem for labels, themes, and layout.

5. **Fix rotated labels with `hjust = 1`.** Rotating alone (`angle = 30`) causes labels to overlap the plot; `hjust = 1` right-aligns them to the tick mark.

6. **`parallelPlot` enables brushing** — clicking an axis variable instantly recolours by value, and selecting a range on any axis filters lines. This makes it far more powerful for interactive discovery.

7. **Add histograms** via `histoVisibility` in `parallelPlot` to understand marginal distributions without switching to a separate plot.

---

## Packages Reference

```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```

| Package | Key Functions |
|---|---|
| `GGally` | `ggparcoord()` |
| `parallelPlot` | `parallelPlot()` |
| `ggplot2` (via tidyverse) | `facet_wrap()`, `theme()`, `element_text()` |
| `readr` (via tidyverse) | `read_csv()` |
| `dplyr` (via tidyverse) | `select()` |

---

## References

- Kam T. S. (2023). [R for Visual Analytics](https://r4va.netlify.app/chap15)
- [`ggparcoord()` documentation](http://ggobi.github.io/ggally/reference/ggparcoord.html) — GGally package
- [`parallelPlot` CRAN documentation](https://cran.r-project.org/web/packages/parallelPlot/parallelPlot.pdf)
- Stephen Few (2006), *Parallel Coordinates: Strengths and Limitations* — Perceptual Edge