---
title: "5e. Treemap Visualisation with R"
author: "Lei Yuhe"
date: "May 14, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview

A **treemap** is a space-filling visualisation that encodes hierarchical data as nested rectangles. Two visual channels are used simultaneously:

- **Rectangle size**: quantitative variable (e.g., total units sold)
- **Rectangle colour**: a second quantitative variable (e.g., median unit price)

Treemaps are especially effective for displaying part-to-whole relationships across multiple levels of a hierarchy, such as property transactions grouped by region planning area, and then project name.

**This chapter covers three approaches:**

| Approach | Package | Function(s) |
|---|---|---|
| Static treemap (full-featured) | `treemap` | `treemap()` |
| Static treemap (ggplot2-based) | `treemapify` | `geom_treemap()` + ggplot2 |
| Interactive treemap | `d3treeR` | `d3tree()` |

---

## Installing and Launching R Packages

```{r}
pacman::p_load(treemap, treemapify, tidyverse)
```

`d3treeR` is not on CRAN and must be installed from GitHub:

```{r}
# install.packages("devtools")          # skip if already installed
library(devtools)
install_github("timelyportfolio/d3treeR")
library(d3treeR)
```

| Package | Role |
|---|---|
| `treemap` | Full-featured static treemaps with 43+ arguments |
| `treemapify` | Treemaps as ggplot2 layers; supports ggplot2 ecosystem |
| `d3treeR` | Interactive, zoomable treemaps built on D3.js |
| `tidyverse` | Data import (`readr`) and wrangling (`dplyr`) |

---

## Data Wrangling

The dataset is `realis2018.csv`, private property transaction records from Singapore's Urban Redevelopment Authority (URA) REALIS portal for 2018.

### Importing the data set

```{r}
realis2018 <- read_csv("data/realis2018.csv")
```

### Data Wrangling and Manipulation

The raw data is at **transaction level** (one row per transaction), which is too disaggregated for a treemap. We need to aggregate it to **project level**, computing:

- `Total Unit Sold` = sum of units per project
- `Total Area` = sum of floor area per project
- `Median Unit Price ($ psm)` = median unit price per project
- `Median Transacted Price` = median transacted price per project

Grouping variables: `Project Name`, `Planning Region`, `Planning Area`, `Property Type`, `Type of Sale`.

Two `dplyr` verbs do this: `group_by()` and `summarise()`.

### Grouped Summaries without the Pipe

```{r}
realis2018_grouped <- group_by(realis2018, `Project Name`,
                               `Planning Region`, `Planning Area`,
                               `Property Type`, `Type of Sale`)

realis2018_summarised <- summarise(realis2018_grouped,
                          `Total Unit Sold`          = sum(`No. of Units`, na.rm = TRUE),
                          `Total Area`               = sum(`Area (sqm)`, na.rm = TRUE),
                          `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
                          `Median Transacted Price`  = median(`Transacted Price ($)`, na.rm = TRUE))
```

> **Note:** `na.rm = TRUE` is required in all aggregation functions (`sum()`, `median()`) because any `NA` in the input would otherwise propagate to the output, silently producing missing aggregates.

This approach names every intermediate object, which gets unwieldy in longer pipelines.

### Grouped Summaries with the Pipe

The pipe operator `%>%` passes the output of one function directly into the next, eliminating intermediate objects:

```{r}
realis2018_summarised <- realis2018 %>%
  group_by(`Project Name`, `Planning Region`,
           `Planning Area`, `Property Type`,
           `Type of Sale`) %>%
  summarise(`Total Unit Sold`           = sum(`No. of Units`, na.rm = TRUE),
            `Total Area`                = sum(`Area (sqm)`, na.rm = TRUE),
            `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
            `Median Transacted Price`   = median(`Transacted Price ($)`, na.rm = TRUE))
```

> **Preferred style:** The pipe version is more readable and idiomatic in tidyverse workflows. Both produce identical results.

---

## Designing Treemap with `treemap` Package

`treemap()` is the core function of the **treemap** package, offering 43+ arguments for fine-grained control.

### Designing a Static Treemap

First, filter to the property type of interest:

```{r}
realis2018_selected <- realis2018_summarised %>%
  filter(`Property Type` == "Condominium", `Type of Sale` == "Resale")
```

---

### Using the Basic Arguments

```{r}
treemap(realis2018_selected,
        index        = c("Planning Region", "Planning Area", "Project Name"),
        vSize        = "Total Unit Sold",
        vColor       = "Median Unit Price ($ psm)",
        title        = "Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)")
```

**Three core arguments:**

| Argument | Purpose | Rule |
|---|---|---|
| `index` | Defines the hierarchy levels | Must have **at least two** column names; first = highest level |
| `vSize` | Maps rectangle size | Column must contain **non-negative** values only |
| `vColor` | Maps rectangle colour | Used together with `type` to determine colour encoding |

> **Common mistake:** Without specifying `type`, `treemap()` defaults to `type = "index"`, which colours rectangles by their hierarchy group rather than by the `vColor` variable's actual values. This produces a misleading result where the rectangles are coloured categorically by planning area, not by unit price intensity.

---

### Working with `vColor` and `type` Arguments

Setting `type = "value"` corrects the colour encoding so rectangles reflect the numeric values of `vColor`:

```{r}
treemap(realis2018_selected,
        index        = c("Planning Region", "Planning Area", "Project Name"),
        vSize        = "Total Unit Sold",
        vColor       = "Median Unit Price ($ psm)",
        type         = "value",
        title        = "Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)")
```

- Rectangles now show different intensities of green reflecting median unit prices.
- The legend is binned into equal-interval bins (e.g., 0–5000, 5000–10000, …).

---

### Colours in `treemap` Package

The `type` argument controls how the `vColor` values are mapped to a colour palette. Two key types:

| `type` | Palette assumption | Colour mapping |
|---|---|---|
| `"value"` | Diverging palette (e.g., RdYlBu) | 0 → midpoint colour; ±max(abs) → ends. Treats the palette as centred at zero. |
| `"manual"` | Any palette | min(values) → left end; max(values) → right end; mean(range) → midpoint. Linear mapping. |

Two relevant arguments work alongside `type`:

- **`palette`**: name of the ColorBrewer (or other) palette to use.
- **`mapping`**: controls the exact value-to-colour mapping (auto-set by `type`).

---

### The `"value"` Type Treemap

```{r}
treemap(realis2018_selected,
        index        = c("Planning Region", "Planning Area", "Project Name"),
        vSize        = "Total Unit Sold",
        vColor       = "Median Unit Price ($ psm)",
        type         = "value",
        palette      = "RdYlBu",
        title        = "Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)")
```

> **Why no red?** The "value" type maps 0 to the middle colour and negative values to the red end of RdYlBu. Since all median unit prices are positive, only the yellow-to-blue half of the palette is used, thus no red appears.

> **Legend range:** Displays only the actual data range (e.g., 5,000–45,000) with pretty rounding, not the full palette range.

---

### The `"manual"` Type Treemap

```{r}
# Avoid: diverging palette with all-positive data is confusing
treemap(realis2018_selected,
        index        = c("Planning Region", "Planning Area", "Project Name"),
        vSize        = "Total Unit Sold",
        vColor       = "Median Unit Price ($ psm)",
        type         = "manual",
        palette      = "RdYlBu",   # problematic for all-positive values
        title        = "Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)")
```

> **Problem:** With `type = "manual"`, the midpoint colour (yellow in RdYlBu) is placed at `mean(range(values))`, and both red and blue ends appear, creating a confusing impression that some values are negative or below a threshold. Use a **sequential palette** instead when all values are on one side.

```{r}
# Correct: sequential palette for all-positive data
treemap(realis2018_selected,
        index        = c("Planning Region", "Planning Area", "Project Name"),
        vSize        = "Total Unit Sold",
        vColor       = "Median Unit Price ($ psm)",
        type         = "manual",
        palette      = "Blues",
        title        = "Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)")
```

> **Palette selection rule:** Use **diverging** palettes (RdYlBu, RdBu) only when data has a meaningful midpoint (e.g., z-scores, change from baseline). Use **sequential** palettes (Blues, Greens, YlOrRd) for all-positive or all-negative data.

---

### Treemap Layout

`treemap()` supports two layout algorithms:

| Algorithm | Reference | Behaviour |
|---|---|---|
| `"squarified"` | Bruls et al. (2000) | Optimises aspect ratios (rectangles close to square); **ignores `sortID`** |
| `"pivotSize"` | Bederson et al. (2002) | Respects `sortID` ordering while maintaining acceptable aspect ratios. **Default.** |

### Working with the `algorithm` Argument

```{r}
treemap(realis2018_selected,
        index        = c("Planning Region", "Planning Area", "Project Name"),
        vSize        = "Total Unit Sold",
        vColor       = "Median Unit Price ($ psm)",
        type         = "manual",
        palette      = "Blues",
        algorithm    = "squarified",
        title        = "Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)")
```

### Using `sortID`

`sortID` only takes effect when `algorithm = "pivotSize"`. It determines the order in which rectangles are placed from **top-left to bottom-right**:

```{r}
treemap(realis2018_selected,
        index        = c("Planning Region", "Planning Area", "Project Name"),
        vSize        = "Total Unit Sold",
        vColor       = "Median Unit Price ($ psm)",
        type         = "manual",
        palette      = "Blues",
        algorithm    = "pivotSize",
        sortID       = "Median Transacted Price",
        title        = "Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)")
```

> Use `sortID` to place the most important or highest-value items in the top-left, where the eye naturally lands first.

---

## Designing Treemap using `treemapify` Package

**`treemapify`** integrates treemaps into **ggplot2** using `geom_treemap()` as a layer. This gives access to the full ggplot2 ecosystem, including scales, themes, labels, and more.

### Designing a Basic Treemap

```{r}
ggplot(data = realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`),
       layout = "scol",
       start  = "bottomleft") +
  geom_treemap() +
  scale_fill_gradient(low = "light blue", high = "blue")
```

**Key aesthetics:**

| Aesthetic | Maps to |
|---|---|
| `area` | Rectangle size (quantitative, non-negative) |
| `fill` | Rectangle colour |

- `layout` and `start` control tile arrangement direction.
- `scale_fill_gradient()` gives direct control over the colour ramp, without the need to choose between `"value"` and `"manual"` types.

---

### Defining Hierarchy

`treemapify` uses `subgroup` and `subgroup2` aesthetics (in `aes()`) to create hierarchical nesting.

**Level 1. Group by Planning Region:**

```{r}
ggplot(data = realis2018_selected,
       aes(area     = `Total Unit Sold`,
           fill     = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`),
       start = "topleft") +
  geom_treemap()
```

**Level 2. Group by Planning Region and Planning Area:**

```{r}
ggplot(data = realis2018_selected,
       aes(area      = `Total Unit Sold`,
           fill      = `Median Unit Price ($ psm)`,
           subgroup  = `Planning Region`,
           subgroup2 = `Planning Area`)) +
  geom_treemap()
```

**Adding boundary lines between hierarchy levels:**

```{r}
ggplot(data = realis2018_selected,
       aes(area      = `Total Unit Sold`,
           fill      = `Median Unit Price ($ psm)`,
           subgroup  = `Planning Region`,
           subgroup2 = `Planning Area`)) +
  geom_treemap() +
  geom_treemap_subgroup2_border(colour = "gray40", size = 2) +
  geom_treemap_subgroup_border(colour  = "gray20")
```

**Border geoms for hierarchy levels:**

| Geom | Draws border around |
|---|---|
| `geom_treemap_subgroup_border()` | Level 1 groups (`subgroup`) |
| `geom_treemap_subgroup2_border()` | Level 2 groups (`subgroup2`) |

> **Tip:** Use darker or thicker borders for higher-level groups (Planning Region) and lighter/thinner borders for lower-level groups (Planning Area) to reinforce the visual hierarchy.

---

## Designing Interactive Treemap using `d3treeR`

`d3treeR` wraps a static `treemap` object in a D3.js-powered interactive widget, adding **drill-down zoom** - clicking a rectangle zooms into that group, revealing its sub-rectangles.

### Designing an Interactive Treemap

The workflow is two steps: build the static treemap, then wrap it with `d3tree()`.

**Step 1 — Build and save the static treemap:**

```{r}
tm <- treemap(realis2018_summarised,
              index        = c("Planning Region", "Planning Area"),
              vSize        = "Total Unit Sold",
              vColor       = "Median Unit Price ($ psm)",
              type         = "value",
              title        = "Private Residential Property Sold, 2017",
              title.legend = "Median Unit Price (S$ per sq. m)")
```

- The treemap object `tm` is saved (not just plotted) by assigning it to a variable.
- Note: `realis2018_summarised` (all property types) is used here, not the filtered `realis2018_selected`.

**Step 2 — Wrap with `d3tree()` for interactivity:**

```{r}
d3tree(tm, rootname = "Singapore")
```

- `rootname`: label shown at the top level before any drill-down.
- **Interaction:** Clicking any rectangle zooms into that group; clicking the breadcrumb trail at the top zooms back out.

> **Key advantage:** `d3treeR` adds drill-down navigation without requiring a different data structure — it reuses the `treemap` object directly.

---

## Comparing the Three Approaches

| Feature | `treemap` | `treemapify` | `d3treeR` |
|---|---|---|---|
| Output | Static plot | Static ggplot2 | Interactive HTML widget |
| Hierarchy levels | Via `index` vector | Via `subgroup` / `subgroup2` | Via `index` (inherited from `treemap`) |
| Colour type control | `type` argument | `scale_fill_*()` functions | Inherited from `treemap` |
| Layout algorithms | squarified, pivotSize | Fixed (squarified default) | Inherited from `treemap` |
| ggplot2 integration | No | Yes (full ecosystem) | No |
| Drill-down zoom | No | No | Yes |
| Border control | Limited | `geom_treemap_*_border()` | Limited |
| Best for | Quick, publication static | Customised ggplot2 figures | Exploratory, interactive |

---

## Key Takeaways

1. **Always specify `type`** in `treemap()`. The default `type = "index"` colours by hierarchy group, not by `vColor` values. 

2. **Match palette type to data.** Use diverging palettes (RdYlBu) only when values span a meaningful zero. For all-positive data (like prices), use sequential palettes (Blues, YlOrRd). Using a diverging palette on one-sided data implies a false midpoint.

3. **`vSize` must be non-negative.** Negative sizes are geometrically undefined for rectangle area, need to check the data before plotting.

4. **`sortID` only works with `algorithm = "pivotSize"`** (the default). If switch to `"squarified"`, sort order is ignored.

5. **`treemapify` is the ggplot2 way.** It's the best choice when you need to combine treemaps with other ggplot2 features, such as faceting, custom themes, additional annotation layers.

6. **`d3treeR` converts any `treemap` object to interactive.** The two-step workflow (build `tm`, then `d3tree(tm)`) means you can prototype statically and interactivise with one extra line.

7. **Hierarchy requires at least two levels** in `treemap()`'s `index`. A single-level index produces a flat (non-nested) treemap with no visible grouping structure.

---

## Packages Reference

```{r}
# Main packages
pacman::p_load(treemap, treemapify, tidyverse)

# d3treeR (GitHub only)
library(devtools)
install_github("timelyportfolio/d3treeR")
library(d3treeR)
```

| Package | Key Functions |
|---|---|
| `treemap` | `treemap()` |
| `treemapify` | `geom_treemap()`, `geom_treemap_subgroup_border()`, `geom_treemap_subgroup2_border()` |
| `d3treeR` | `d3tree()` |
| `dplyr` (tidyverse) | `group_by()`, `summarise()`, `filter()` |
| `readr` (tidyverse) | `read_csv()` |
| `ggplot2` (tidyverse) | `ggplot()`, `aes()`, `scale_fill_gradient()` |